When Subgraph Isomorphism is Really Hard, and Why This Matters for Graph Databases

### 2017-11-01 paper/paper.tex
More of the things

    \caption{Cumulative number of instances solved as a function of search space size, using four
    graph database datasets and a simple domain-based subgraph isomorphism algorithm. Note that the
    $x$-axis shows recursive calls, and does \emph{not} use a log scale.}
    \label{figure:biiiig-data}
\end{figure}

?? Redo these using CP.

Rather than the modern techniques discussed used by Glasgow and LAD, we deliberately select a very
simple starting point: a forward-checking algorithm which uses the ``smallest domain first''
heuristic, the simple counting all-different propagator, and degree and neighbourhood degree
sequence filtering at the top of search. Even this simple algorithm finds every instance nearly
trivial, and no exponential behaviour is observed. As we show in \cref{figure:biiiig-data}, the
\emph{hardest} instance for the AIDS dataset requires 59 recursive calls, and 146,175 of the
instances can be solved without search.  For PDBS, the hardest problem requires 1,776 recursive
calls (which occurs 20 times, due to duplicated queries), and none of the rest require more than 182
calls. For PCM the hardest problem requires 3,261 calls, and for PPI, 26. In other words, none of
these instances are in any way computationally hard even for a simple domain-based algorithm, even
before introducing stronger filtering, inference, or heuristics. Furthermore it is certainly not the
case that a sequential scan is infeasible as so many filter~/ verify papers claim.  The exponential
behaviour seen by others in these instances is purely down to a bad choice of subgraph isomorphism
algorithm, and these instances are not inherently ``really hard''.

@@

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we had not used a ``smallest domain first'' heuristic, and instead used only
degree-based heuristics explained earlier in this paper? Although we still find all four of these
datasets easy, and the changes to the runtimes are hard to detect (more time is spent in
initialisation than in search), the worst case in the AIDS dataset now takes 4,051 recursive calls
as opposed to 59.  How about if, additionally, we do not detect domain wipeouts until attempting to
branch on an empty variable?  Suddenly three of the four datasets become extremely hard, with many
instances now timing out after making hundreds of millions of recursive calls; the PPI dataset now
requires up to 168,451 calls for some queries. Furthermore, slight changes to the input vertex
ordering can now make some of these ``hard'' instances easy again.

This backs up our suspicions: the apparent success of \citet{DBLP:conf/edbt/KatsarouNT17}'s
technique is due to the ease of sometimes getting better results out of a poor algorithm, rather
than being an inherent hardness property, and it would not help on really hard problem instances.

>>>>

    \caption{Cumulative number of instances solved as a function of search space size, using four
    graph database datasets and a simple constraint programming subgraph isomorphism algorithm. Note
    that the $x$-axis shows recursive calls, and does \emph{not} use a log scale.}
    \label{figure:biiiig-data}
\end{figure}

Rather than the modern techniques used by Glasgow and LAD, we deliberately select a very simple
starting point: a constraint programming model implemented in Gecode 5.1.0 using only
toolkit-provided constraints and heuristics. The model uses a ``smallest domain first'' variable
ordering heuristic with tie-breaking on degree, an all-different constraint, extensional constraints
for adjacency, and labelled-degree filtering at the top of search. Even this simple approach finds
nearly every instance trivial, and no exponential behaviour is observed. As we show in
\cref{figure:biiiig-data}, the \emph{hardest} instance for the AIDS dataset requires 287 recursive
calls, and 149,268 of the instances can be solved without search.  For PDBS, the hardest problem
requires 3,161 recursive calls (which occurs 20 times, due to duplicated queries). For both PCM and
PPI, the hardest problem requires 23 recursive calls.  In other words, none of these instances are
in any way computationally hard even for a simple domain-based algorithm, even before introducing
stronger filtering, inference, or heuristics.  Furthermore it is certainly not the case that a
sequential scan is infeasible as so many filter~/ verify papers claim.  The exponential behaviour
seen by others in these instances is purely down to a bad choice of subgraph isomorphism algorithm,
and these instances are not inherently ``really hard''.

@@

To test this suggestion, we return briefly to the experiments on the datasets discussed earlier in
this section. What if we modify our solver so that it does not use ``smallest domain first'', and
does not detect domain wipeouts until attempting to branch on an empty variable?  Suddenly three of
the four datasets become extremely hard, with many instances now timing out after making hundreds of
millions of recursive calls, and the PPI dataset now requires hundreds of thousands of calls for
some queries.  Furthermore, slight changes to the input vertex ordering can now make some of these
``hard'' instances easy again.  This backs up our suspicions: the apparent success of
\citet{DBLP:conf/edbt/KatsarouNT17}'s technique is due to the ease of sometimes getting better
results out of a poor algorithm, rather than being an inherent hardness property, and it would not
help on really hard problem instances.
